Constructing Sailing Match Race Schedules: Round Robin Pairing Lists

### 2015-04-07 paper.tex
Indenting
Automaton, criterion

\begin{framed}
Principal Criteria in Order of Priority:
\begin{enumerate}
\item Each skipper sails against each other skipper once.
\item *When skippers have an even number of matches, they have the same number of port and starboard
    assignments.
\item *When skippers have an odd number of matches, the first half of the skippers will have one
    more starboard assignment.
\item No skipper in the last match of a flight should be in the first match of the next flight.
\item No skipper should have more than two consecutive port or starboard assignments.
\item Each skipper should be assigned to match 1, match 2, etc. in a flight as equally as possible.
\item In flights with five or more matches, no skipper should be in the next-to-last match in a
    flight and then in the first match of the next flight.
\item If possible, a skipper should be starboard when meeting the nearest lowest ranked skipper
    (i.e.\ \#1 will be starboard against \#2, \#3 will be starboard against \#4).
\item Close-ranked skippers meet in the last flight.
\item Minimise the number of boat changes.
\item Skippers in the last match of a flight do not change boats.
\item Skippers in new boats do not sail in the first match of the next flight.
\item Skippers have a reasonable sequence of matches and blanks.
\end{enumerate}
Note that criteria 10, 11 and 12 only apply when there are fewer boats than skippers; 11 and 12
override 6 when changes are required.
\end{framed}

We have rephrased criteria 2 \& 3 (denoted *) to clarify an error in \cite{isaf}. Note that the
order of matches within a flight is significant and is constrained, as is the order of flights
within a schedule and the position of skippers within a match. This permits fair schedules that
provide the sufficient changeover time for skippers changing boats, etc.

Criteria 12 defines what is meant by a boat change: if in flight $i$ this skipper is not in a match
(i.e. it is a bye for this skipper) but is in a match in flight $i+1$ then he has to get into a
boat, rearrange all his kit and set the boat to his preference before racing commences, and this is
a boat change.

Next, we note that skippers can be ranked prior to an event, based on their performance in
previous events\footnote{\url{http://www.sailing.org/rankings/match/}}, which is used to seed the
skippers. The ordering of skippers within a match signifies their starting position (port or
starboard), as skippers allocated the starboard starting position gain a small competitive
advantage in that match---criteria 8 accomplishes the seedings.

Criteria 10 is ambiguous: this could be interpreted on a schedule basis: i.e.\ to minimise the
overall number of changes in the entire pairing list schedule; or alternatively as well as a fair,
but minimal number of changes across all skippers. In our work, we take the latter viewpoint.

There are some conflicts inherent in the rules. Take criteria 4 and 11, assume we have 4 boats, and
in flight $i$ the last match is the pair of skippers ($x,y$). Criteria 11 dictates that skippers $x$
and $y$ must appear in the next flight, $i+1$, and criteria 4 that neither can be first in the next
flight. This forces skippers $x$ and $y$ to compete in flight $i+1$ as the last match, violating
criteria 1. Therefore, although not explicitly stated, there are no legal schedules with less than 6
boats.

@@

The schedules are produced in four stages. The first stage produces a schedule that respects
criteria 1, 4, 11 and 12, and minimises criteria 10 (boat changes). The second stage constructs a
schedule that respects criteria 1, 4, 11, 12 (again), has criteria 10 as a hard constraint generated
from first stage, and minimises criteria 6 (balance). This results in a schedule that attempts to
minimise boat changes \emph{and} balance. The third stage satisfies criteria 9, and is a simple
translation of the schedule produced in stage 2.  The final stage orients skippers within matches to
address criteria 2, 3, 5 and 8.

@@

\paragraph{State transitions:} The $\state$ variables are then used to impose match race criteria 4
(if last in flight $i$ then not first in flight $i+1$), criteria 11 (if last in flight $i$ then not
in a bye in flight $i+1$) and criteria 12 (if flight $i$ is a bye then flight $i$ is not first).
These criteria are imposed by \eqref{C3}, which is a deterministic finite automata (DFA) constraint
\cite{Pesant04}. The transitions for the DFA are shown pictorially in Figure~\ref{skipper5}: arcs
represent a transition from one state to another. The DFA is encoded as a set of transition objects
$\langle q_{i},\iota,q_{j} \rangle$, where a transition is made from state $q_{i}$ to state $q_{j}$
when encountering input $\iota$. In addition we specify the accepting states, which are all states
except the start state (white) and the bye state (pink).  This constraint also ensures that if a
skipper has finished his matches in flight $i$ (i.e.\ $\END$) then he has finished his matches in
flight $i+1$ (and so on).

@@

    \end{tikzpicture}
    \caption{A pictorial representation of a skipper (skipper 5) with multicoloured state and grey
        timeslots. The schedule for skipper 5 is in bold and the deterministic finite automata for
        criteria 4, 11 and 12 is drawn with state $\START$ in white, $\FIRST$ in blue, $\MID$ in
        yellow, $\LAST$ in green, $\BYE$ in pink and $\END$ in red.}\label{skipper5}
\end{figure}

\paragraph{Match perspective:} We are now in a position to model criteria 1 (each skipper sails
against every other skipper) and optimisation criteria 10 (minimise boat changes).  In the second
half of Figure~\ref{model:stage1} we present a match perspective of the schedule, using a two
dimensional array of variables $\match$ \eqref{V5}, where $\match[{i,j}]$ is the time slot that
skippers $\sigma[{i}]$ and $\sigma[{j}]$ meet in a match. Only the half above the diagonal is
represented, and the lower half is made up of exactly the same variables, i.e.\ $\match[{i,j}]$ is
\emph{exactly} the same constrained variable as $\match[{j,i}]$. Constraint \eqref{C6} states that
a match between skippers $\sigma[{i}]$ and $\sigma[{j}]$ takes place at time $k$ (i.e.\
$\match[{i,j}] = k$) if and only if  skipper $\sigma[{i}]$'s $k^{th}$ time slot is skipper
$\sigma[{j}]$ and conversely that skipper $\sigma[{j}]$'s $k^{th}$ time slot is skipper
$\sigma[{i}]$). Variables \eqref{V6} and constraints \eqref{C7} then convert this from time slots to
flights, i.e.\ $\modMatch[{i,j}]$ is the flight in which skippers $\sigma[{i}]$ and $\sigma[{j}]$
meet for a match. Finally, constraint \eqref{C8} ensures that each skipper's match occurs in
different flights. Also, since $\modMatch[{i,j}] \equiv \modMatch[{j,i}]$, setting rows to be all
different also ensures that columns are all different.

@@

\paragraph{Optimisation criteria:} Finally we have the optimisation criteria (criteria 10) to
minimise the number of boat changes. This is handled by \eqref{V8} and \eqref{C10}.

@@

Having produced a schedule that minimises boat changes, we then minimise criteria 6, imbalance, by
extending the model as in Figure~\ref{model:stage2}.  Assuming the first stage produced a schedule
with $\beta$ boat changes we post this as a hard constraint \eqref{C11} in stage 2. For each skipper
$\sigma$ we introduce a zero/one variable $position[{i,j}]$ \eqref{V9} which is 1 if and only if
skipper $\sigma$ is in a match in time $m \cdot j + i$, i.e.\ in the $i^{th}$ match of the  $j^{th}$
flight \eqref{C12}. Imbalance \eqref{V10} is then computed for each position, as the absolute amount
of over or under expected presence in a given position \eqref{C13}. Finally, we compute the maximum
imbalance over all positions for this skipper (\ref{V11}, \ref{C14}).

@@

Stage 3 then takes as input the schedule produced in stage 2 and renames skippers in the last round
of the last flight to be $(0,1)$, satisfying criteria 9. This is a simple linear process. The
schedule produced for 7 skippers and 6 boats is shown in Table~\ref{tab2}.

@@

The final stage, stage 4, orients skippers within matches, such that they are either port (red) or
starboard (green). This is done to satisfy criteria 2 and 3, 5 and 8. A sketch of the constraint
model is given with a supporting diagram, Figure~\ref{oriented}.  A sequence of zero/one constrained
integer variables, of length $n-1$, is produced for each skipper. These correspond to the sequence
of skippers they meet in matches. For example, taking the schedule in Table~\ref{tab2}, skipper
$\sigma[{2}]$ meets skippers in order: 3,0,1,4,5,6, and skipper $\sigma[{5}]$ in the order
4,6,0,1,3,2.  Therefore, skipper $\sigma[{2}]$'s 4$^{th}$ variable must take a different value from
$\sigma[{5}]$ 5$^{th}$ variable. There are summation constraints on each skipper to enforce criteria
2 and 3 (equal number of port and starboard matches). Criteria 8 is enforced by posting constraints
such that in a match ($\sigma[{i}],\sigma[{j}]$) where $\abs*{i - j} = 1$, the higher indexed
skipper (lower ranked) is on the port side. A DFA constraint is posted for criteria 5 (restricting
sequences of port and starboard assignments).

@@

\begin{figure}[tb]
\centering
\includegraphics[height=10.2cm,width=13.2cm]{oriented.pdf}
\caption{A pictorial representation of the orientation process, stage 4. Top right is the actual
oriented schedule. Below that is the dfa to satisfy criteria 5. On the left is the zero/one
variables, a row representing a skippers sequence of competitors. An edge between variables
corresponds to a matched pair, that must take different values. Bottom right is a sketch by third
author.}\label{oriented}
\end{figure}

@@

\paragraph{8 skippers and 6 boats:} We first analyse the ISAF schedule, Table~\ref{08-06} on the
left. Criteria 4 (skippers in last match in a flight cannot be first in next flight) is violated on
3 occasions (skipper 1 in flight 3, skipper 7 in flight 4, skipper 0 in flight 7). Criteria 5 (no
more than two consecutive port or starboard assignments) is violated for skipper 6 (flights 1 to 3)
and 7 (flights 7 to 9). Criteria 6 (imbalance) is violated for skippers 6 and 8. Criteria 12  (new
boats do not sail first) is violated in flight 9 for skipper 5. Finally, the schedule has 8 boat
changes. Our schedule, Table~\ref{08-06} on the right, respects all criteria and has 6 boat changes
and a maximum imbalance of 1.

@@

\paragraph{10 skippers and 6 boats:} Table \ref{10-06} shows the ISAF schedule on the right for 10
skippers and 6 boats and on the right, our schedule.  The ISAF schedule violates criteria 5 (no more
than 2 consecutive port or starboard assignments) 10 times (twice for skipper 0, 1 and 6 and once
for skippers 3, 7, 8 and 9).  Criteria 12 (new boats do not sail first) is violated 7 times (in
flight 3 for skippers 6 and 2, flight 5 for skipper 8, flight 6 for skipper 7, flight 8 for skipper
1, flight 9 for skipper 4 and flight 12 for skipper 4 again). There are 22 boat changes, with the
minimum changes for a skipper being 1 and the maximum 3. Our schedule, right of Table \ref{10-06}
satisfies all criteria, has 12 boat changes with the minimum changes for a skipper being 0 and the
maximum 2, and a maximum imbalance of 1.

@@

We have only reported a handful of our schedules, however there are missing schedules. Examples of
these missing schedules are 10 skippers and 8 boats, 11 skippers with less than 10 boats, 12
skippers and 8 boats, 13 skippers and 8 boats and 14 skippers with less than 10 boats. None of these
schedules have been published by the ISAF, although we expect that they can be produced by
selectively violating some of constraints. We have also not yet encoded criteria 7, addressing the
situation of 10 or more boats. To do this we will have to modify the DFA constraint used in the
first two stages.

>>>>

\begin{framed}
\noindent Principal Criteria in Order of Priority:
\begin{enumerate}
\item Each skipper sails against each other skipper once.
\item *When skippers have an even number of matches, they have the same number of port and starboard
    assignments.
\item *When skippers have an odd number of matches, the first half of the skippers will have one
    more starboard assignment.
\item No skipper in the last match of a flight should be in the first match of the next flight.
\item No skipper should have more than two consecutive port or starboard assignments.
\item Each skipper should be assigned to match 1, match 2, etc. in a flight as equally as possible.
\item In flights with five or more matches, no skipper should be in the next-to-last match in a
    flight and then in the first match of the next flight.
\item If possible, a skipper should be starboard when meeting the nearest lowest ranked skipper
    (i.e.\ \#1 will be starboard against \#2, \#3 will be starboard against \#4).
\item Close-ranked skippers meet in the last flight.
\item Minimise the number of boat changes.
\item Skippers in the last match of a flight do not change boats.
\item Skippers in new boats do not sail in the first match of the next flight.
\item Skippers have a reasonable sequence of matches and blanks.
\end{enumerate}
Note that criteria~10, 11 and~12 only apply when there are fewer boats than skippers; 11 and~12
override~6 when changes are required.
\end{framed}

We have rephrased criteria~2 and~3 (denoted *) to clarify an error in \cite{isaf}. Note that the
order of matches within a flight is significant and is constrained, as is the order of flights
within a schedule and the position of skippers within a match. This permits fair schedules that
provide the sufficient changeover time for skippers changing boats, etc.

Criterion~12 defines what is meant by a boat change: if in flight $i$ this skipper is not in a match
(i.e. it is a bye for this skipper) but is in a match in flight $i+1$ then he has to get into a
boat, rearrange all his kit and set the boat to his preference before racing commences, and this is
a boat change.

Next, we note that skippers can be ranked prior to an event, based on their performance in
previous events\footnote{\url{http://www.sailing.org/rankings/match/}}, which is used to seed the
skippers. The ordering of skippers within a match signifies their starting position (port or
starboard), as skippers allocated the starboard starting position gain a small competitive
advantage in that match---criterion~8 accomplishes the seedings.

Criterion~10 is ambiguous: this could be interpreted on a schedule basis: i.e.\ to minimise the
overall number of changes in the entire pairing list schedule; or alternatively as well as a fair,
but minimal number of changes across all skippers. In our work, we take the latter viewpoint.

There are some conflicts inherent in the rules. Take criteria~4 and~11, assume we have 4 boats, and
in flight $i$ the last match is the pair of skippers ($x,y$). Criterion~11 dictates that skippers $x$
and $y$ must appear in the next flight, $i+1$, and criterion~4 that neither can be first in the next
flight. This forces skippers $x$ and $y$ to compete in flight $i+1$ as the last match, violating
criterion~1. Therefore, although not explicitly stated, there are no legal schedules with less than 6
boats.

@@

The schedules are produced in four stages. The first stage produces a schedule that respects
criteria~1, 4, 11 and~12, and minimises criteria~10 (boat changes). The second stage constructs a
schedule that respects criteria~1, 4, 11, 12 (again), has criterion~10 as a hard constraint generated
from first stage, and minimises criterion~6 (balance). This results in a schedule that attempts to
minimise boat changes \emph{and} balance. The third stage satisfies criterion~9, and is a simple
translation of the schedule produced in stage 2.  The final stage orients skippers within matches to
address criteria~2, 3, 5 and~8.

@@

\paragraph{State transitions:} The $\state$ variables are then used to impose match race criterion~4
(if last in flight $i$ then not first in flight $i+1$), criterion~11 (if last in flight $i$ then not
in a bye in flight $i+1$) and criterion~12 (if flight $i$ is a bye then flight $i$ is not first).
These criteria are imposed by \eqref{C3}, which is a deterministic finite automaton (DFA) constraint
\cite{Pesant04}. The transitions for the DFA are shown pictorially in Figure~\ref{skipper5}: arcs
represent a transition from one state to another. The DFA is encoded as a set of transition objects
$\langle q_{i},\iota,q_{j} \rangle$, where a transition is made from state $q_{i}$ to state $q_{j}$
when encountering input $\iota$. In addition we specify the accepting states, which are all states
except the start state (white) and the bye state (pink).  This constraint also ensures that if a
skipper has finished his matches in flight $i$ (i.e.\ $\END$) then he has finished his matches in
flight $i+1$ (and so on).

@@

    \end{tikzpicture}
    \caption{A pictorial representation of a skipper (skipper 5) with multicoloured state and grey
        timeslots. The schedule for skipper 5 is in bold and the deterministic finite automaton for
        criteria~4, 11 and~12 is drawn with state $\START$ in white, $\FIRST$ in blue, $\MID$ in
        yellow, $\LAST$ in green, $\BYE$ in pink and $\END$ in red.}\label{skipper5}
\end{figure}

\paragraph{Match perspective:} We are now in a position to model criterion~1 (each skipper sails
against every other skipper) and optimisation criterion~10 (minimise boat changes).  In the second
half of Figure~\ref{model:stage1} we present a match perspective of the schedule, using a two
dimensional array of variables $\match$ \eqref{V5}, where $\match[{i,j}]$ is the time slot that
skippers $\sigma[{i}]$ and $\sigma[{j}]$ meet in a match. Only the half above the diagonal is
represented, and the lower half is made up of exactly the same variables, i.e.\ $\match[{i,j}]$ is
\emph{exactly} the same constrained variable as $\match[{j,i}]$. Constraint \eqref{C6} states that
a match between skippers $\sigma[{i}]$ and $\sigma[{j}]$ takes place at time $k$ (i.e.\
$\match[{i,j}] = k$) if and only if  skipper $\sigma[{i}]$'s $k^{th}$ time slot is skipper
$\sigma[{j}]$ and conversely that skipper $\sigma[{j}]$'s $k^{th}$ time slot is skipper
$\sigma[{i}]$). Variables \eqref{V6} and constraints \eqref{C7} then convert this from time slots to
flights, i.e.\ $\modMatch[{i,j}]$ is the flight in which skippers $\sigma[{i}]$ and $\sigma[{j}]$
meet for a match. Finally, constraint \eqref{C8} ensures that each skipper's match occurs in
different flights. Also, since $\modMatch[{i,j}] \equiv \modMatch[{j,i}]$, setting rows to be all
different also ensures that columns are all different.

@@

\paragraph{Optimisation criteria:} Finally we have the optimisation criterion (criterion~10) to
minimise the number of boat changes. This is handled by \eqref{V8} and \eqref{C10}.

@@

Having produced a schedule that minimises boat changes, we then minimise criterion~6, imbalance, by
extending the model as in Figure~\ref{model:stage2}.  Assuming the first stage produced a schedule
with $\beta$ boat changes we post this as a hard constraint \eqref{C11} in stage 2. For each skipper
$\sigma$ we introduce a zero/one variable $position[{i,j}]$ \eqref{V9} which is 1 if and only if
skipper $\sigma$ is in a match in time $m \cdot j + i$, i.e.\ in the $i^{th}$ match of the  $j^{th}$
flight \eqref{C12}. Imbalance \eqref{V10} is then computed for each position, as the absolute amount
of over or under expected presence in a given position \eqref{C13}. Finally, we compute the maximum
imbalance over all positions for this skipper (\ref{V11}, \ref{C14}).

@@

Stage 3 then takes as input the schedule produced in stage 2 and renames skippers in the last round
of the last flight to be $(0,1)$, satisfying criterion~9. This is a simple linear process. The
schedule produced for 7 skippers and 6 boats is shown in Table~\ref{tab2}.

@@

The final stage, stage 4, orients skippers within matches, such that they are either port (red) or
starboard (green). This is done to satisfy criteria~2 and~3, 5 and~8. A sketch of the constraint
model is given with a supporting diagram, Figure~\ref{oriented}.  A sequence of zero/one constrained
integer variables, of length $n-1$, is produced for each skipper. These correspond to the sequence
of skippers they meet in matches. For example, taking the schedule in Table~\ref{tab2}, skipper
$\sigma[{2}]$ meets skippers in order: 3,0,1,4,5,6, and skipper $\sigma[{5}]$ in the order
4,6,0,1,3,2.  Therefore, skipper $\sigma[{2}]$'s 4$^{th}$ variable must take a different value from
$\sigma[{5}]$ 5$^{th}$ variable. There are summation constraints on each skipper to enforce
criteria~2 and~3 (equal number of port and starboard matches). Criterion~8 is enforced by posting
constraints such that in a match ($\sigma[{i}],\sigma[{j}]$) where $\abs*{i - j} = 1$, the higher
indexed skipper (lower ranked) is on the port side. A DFA constraint is posted for criterion~5
(restricting sequences of port and starboard assignments).

@@

\begin{figure}[tb]
\centering
\includegraphics[height=10.2cm,width=13.2cm]{oriented.pdf}
\caption{A pictorial representation of the orientation process, stage 4. Top right is the actual
oriented schedule. Below that is the DFA to satisfy criterion~5. On the left is the zero/one
variables, a row representing a skippers sequence of competitors. An edge between variables
corresponds to a matched pair, that must take different values. Bottom right is a sketch by third
author.}\label{oriented}
\end{figure}

@@

\paragraph{8 skippers and 6 boats:} We first analyse the ISAF schedule, Table~\ref{08-06} on the
left. Criterion~4 (skippers in last match in a flight cannot be first in next flight) is violated on
3 occasions (skipper 1 in flight 3, skipper 7 in flight 4, skipper 0 in flight 7). Criterion~5 (no
more than two consecutive port or starboard assignments) is violated for skipper 6 (flights 1 to 3)
and 7 (flights 7 to 9). Criterion~6 (imbalance) is violated for skippers 6 and 8. Criterion~12  (new
boats do not sail first) is violated in flight 9 for skipper 5. Finally, the schedule has 8 boat
changes. Our schedule, Table~\ref{08-06} on the right, respects all criteria and has 6 boat changes
and a maximum imbalance of 1.

@@

\paragraph{10 skippers and 6 boats:} Table \ref{10-06} shows the ISAF schedule on the right for 10
skippers and 6 boats and on the right, our schedule.  The ISAF schedule violates criterion~5 (no more
than 2 consecutive port or starboard assignments) 10 times (twice for skipper 0, 1 and 6 and once
for skippers 3, 7, 8 and 9).  Criterion~12 (new boats do not sail first) is violated 7 times (in
flight 3 for skippers 6 and 2, flight 5 for skipper 8, flight 6 for skipper 7, flight 8 for skipper
1, flight 9 for skipper 4 and flight 12 for skipper 4 again). There are 22 boat changes, with the
minimum changes for a skipper being 1 and the maximum 3. Our schedule, right of Table \ref{10-06}
satisfies all criteria, has 12 boat changes with the minimum changes for a skipper being 0 and the
maximum 2, and a maximum imbalance of 1.

@@

We have only reported a handful of our schedules, however there are missing schedules. Examples of
these missing schedules are 10 skippers and 8 boats, 11 skippers with less than 10 boats, 12
skippers and 8 boats, 13 skippers and 8 boats and 14 skippers with less than 10 boats. None of these
schedules have been published by the ISAF, although we expect that they can be produced by
selectively violating some of constraints. We have also not yet encoded criterion~7, addressing the
situation of 10 or more boats. To do this we will have to modify the DFA constraint used in the
first two stages.
